Plant Modeling Using Environmental Parameters

### 2014-03-28 doc/paper.tex
Finish initial paper draft

\usepackage{mdwlist}
\usepackage{graphics}
\usepackage{graphicx}
\usepackage{listings}
\usepackage{enumitem}
\usepackage[utf8]{inputenc}

@@

The L-Systems are implemented in the LSystem and LRule classes, which are
responsible for storing the production rules associated with each plant and
iterating the plants phenotype as the plant grows throughout the simulation. The
Gene and Plant classes store representations of each living plant. The Gene
class represents the plant genotype as a combination of the L-System production
rules along with additional information such as stem and leaf size, stem and
flower colors, the number of seeds produced, and the age at which the plant
matures. Each individual plants phenotype is represented using the Plant class
and contains it's X/Y position and orientation inside the world, along with the
current age and a L-System string (L-String) representing the plants current
shape. The Plant class is also responsible for calculating each individual
plants fitness values at the end of each simulation step.

@@

@@

There are three basic environmental parameters which has an effect on the
plants. The amount of sun (0-100\%), water (0-100\%) and the nutrition available
in the ground.

@@

>>>>

\usepackage{mdwlist}
\usepackage{graphics}
\usepackage{graphicx}
\usepackage{listings}
\usepackage{enumitem}
\usepackage{amsmath}
\usepackage[utf8]{inputenc}

@@

The L-Systems are implemented in the LSystem and LRule classes, which are
responsible for storing the production rules associated with each plant and
iterating the plants phenotype as the plant grows throughout the simulation. The
Gene and Plant classes store representations of each living plant. The Gene
class represents the plant genotype as a combination of the L-System production
rules along with additional information such as stem and leaf size, stem and
flower colors, the number of seeds produced, and the age at which the plant
matures. Each individual plants phenotype is represented using the Plant class
and contains it's X/Y position and orientation inside the world, along with the
current age and a L-System string (L-String) representing the plants current
shape. The Plant class is also responsible for calculating each individual
plants fitness values at the end of each simulation tick.

@@

At the start of each simulation tick, each existing plant updates a fitness
value based on the current environmental parameters and the plants current
growth state. The list of plants is then sorted based on fitness and unfit
plants are removed if the environment is over crowded.

@@

There are three basic environmental parameters which have an effect on the
plants.

\begin{description}[leftmargin=!,labelindent=0.2in,labelwidth=0.5in]
  \item[sun]       The amount of sunlight
  \item[water]     The availability of water
  \item[nutrition] The nutritional content of the soil
\end{description}

Each of these parameters initially ranges from 0 to 100 and can be changed by
the user interacting with the software. In addition, a mutation parameter
controls the rate of change in the plant genomes.

For the purpose of fitness calculations, a ``genetic pressure'' version of each
of these parameters is calculated and used in the fitness function.
\[pressure = 1 - (parameter / 100)\]

\subsubsection{Plant parameters}

When calculating each plants fitness several plant parameters are used:

\begin{description}[leftmargin=!,labelindent=0.2in,labelwidth=0.1in]
  \item[Stem ratio]    \hfill \\
    Normalized stem width, $0 .. 1$
  \item[Leaf ratio]    \hfill \\
    Normalized leaf size, $0 .. 1$
  \item[Stem count]    \hfill \\
    The number of stem segments the plant has grown thus far.
  \item[Leaf count]    \hfill \\
    The number of leaves grown thus far.
  \item[Storage space] \hfill \\
    The total storage space available to the plant. We do not calculate an
    actual volume but instead just use a combination of the stem width from the
    plants genes and the number of stem segments.
    \[ storageSpace = stemRatio \times stemCount \]
  \item[Surface area]  \hfill \\
    The total surface area available in the leaves for photosynthesis. Like the
    storage space, this is a combination of the leaf size and the number of
    leaves.
    \[ surfaceArea = leafRatio \times leafCount \]
\end{description}

\subsubsection{Fitness calculation}

The plant fitness function is divided by water, sun, an nutrient fitness. Each
of these is calculated separately and the total fitness is the sum. We
implemented the fitness equations such that if all the environmental parameters
are equal, then all the fitness values will also be equal and the plants will
evolve randomly. In a more comprehensive system, some plants could be considered
more or less fit in any environment including an environment with equal
environmental parameters.

The following guidelines were used by our fitness function:

\paragraph{Water fitness}
\begin{itemize}
  \item Large stems (high storage space) can conserve water and result in higher
    fitness in dry environments.
  \item A large surface area results in larger water loss and a lower
    fitness in dry environments.
\end{itemize}
\vspace{-0.5em}
\begin{eqnarray*}
waterFitness &=& (storageSpace \times waterPressure) \\
             &-& (surfaceArea \times waterPressure)
\end{eqnarray*}

\paragraph{Sunlight fitness}
\begin{itemize}
  \item A large surface area collects more light and results in a higher fitness
    in shady environments.
  \item More stem segments will result in a taller plant which can college more
    light. We do not calculate actual heights, but assume that stem segments
    improve light gathering.
\end{itemize}
\vspace{-0.5em}
\begin{eqnarray*}
sunFitness &=& (2 \times surfaceArea \times sunPressure) \\
           &+& (stemCount \times sunPressure)
\end{eqnarray*}

\paragraph{Nutrition fitness}
\begin{itemize}
  \item Larger stems, more stem segments, and a larger surface all result in
    higher nutrient requirements and thus lower the fitness in low quality soil.
\end{itemize}
\vspace{-0.5em}
\begin{eqnarray*}
nutrientFitness &=&-(surfaceArea \times nutrientPressure) \\
                &+&-(storageSpace \times nutrientPressure) \\
                &+&-(storageSpace \times nutrientPressure)
\end{eqnarray*}

@@

During our simulations we were able to adjust the environmental parameters in
order to control the growth and evolution of different plant species. In many
cases the evolved plants share some resemblance with natural plants found in
similar environments. Some example of this include stout leafless plants
resembling cacti which evolve in dry environments, large leafy plants resembling
trees that evolve in shady environments, and small shrubby or grass like plants
that evolve in poor soil conditions.

In future programs we would recommend improving the genetic fitness functions by
taking into consideration environmental information related to the neighboring
plants in the environment. This would use more advanced lighting and collision
detection to calculate sunlight based on shadows in the environment.

In addition, more realistic fitness levels could be used that are not linear
with regards to the amount of pressure from different environmental parameters.
In our system, a leafy plant may always win over other plants in a shady
simulation environment even though such as large pant may not be support in a
physical environment to do water or soil limitations.

Some difficulty was encountered evolving plants to appear realistic, and many
limitations are inherit which are difficult to simulation in a computer graphics
setting. For example, two physical plants which share a remarkable resemblances
my be suited to different environments due to small evolutionary changes. Waxy
leaves, for example, could result in a leafy plant well suited to a dry
environment.

Overall, we have shown that artificial plants based on L-Systems can be evolved
in realistic ways based on abstract environmental parameters. Using L-Strings to
represent plants results in both realistic looking shapes and provides the
foundation for genetic algorithm for controlling the development of plant
species.
